### 2016-02-18 stitched.tex
ENH: final
ENH:  more.

\usepackage{booktabs}
\usepackage[deletedmarkup=xout]{changes}
\usepackage[font={small},labelfont=bf,labelsep=colon]{caption}
\linespread{1.5}
\usepackage[compact]{titlesec}
\usepackage{enumitem}
\usepackage{tikz}
\def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
\setlist{nolistsep}
\titlespacing{\section}{2pt}{*0}{*0}
\titlespacing{\subsection}{2pt}{*0}{*0}
\titlespacing{\subsubsection}{2pt}{*0}{*0}
\setlength{\parskip}{3pt}
\setremarkmarkup{(#2)}
\definechangesauthor[name={Nick Tustison}, color=red]{nt}
\definechangesauthor[name={James Stone}, color=magenta]{js}
\definechangesauthor[name={Lisa Wilde}, color=cyan]{lw}
\definechangesauthor[name={Andy Mayer}, color=green]{am}
\definechangesauthor[name={Harvey Levin}, color=blue]{hl}
\definechangesauthor[name={Brian Taylor}, color=purple]{bt}
\definechangesauthor[name={David Tate}, color=orange]{dt}

@@

@@

White matter hyperintensities (WMHs) are foci of abnormally increased
signal intensity seen within white matter regions within the cerebrum
and brainstem on fluid attenuation inversion recovery (FLAIR) magnetic
resonance imaging (MRI) sequences. These lesions are distinguished from
prominent perivascular spaces seen on T2-weighted imaging by the lack of
fluid suppression of signal on FLAIR sequences. WMHs in the
periventricular and deep brain regions are associated with normal aging
and neurological conditions including hypertension and stroke. WMHs are
also a frequent finding following traumatic brain injury (TBI) and have
been correlated with functional outcome and injury severity in both
pediatric {[}1, 2{]} and adult {[}3--6{]}. Further, the regional
distribution and volume of WMHs have been shown to possess prognostic
value in the TBI patient {[}2, 6--8{]}. Specifically, lesion volume in
corpus callosum correlates with functional scores in the acute phase
following injury, while lesion volume in frontal lobes correlates with
scores at 1 year following injury {[}6{]}. Further, volume of FLAIR
lesions within the corpus callosum, brainstem, and thalamus in patients
with severe TBI correlates with Glasgow Outcome-Extended (GOS-E) scores
{[}4{]}. Additionally, the regional distribution of FLAIR lesions within
the pons, midbrain, hypothalamus, basal forebrain, parietal, temporal,
occipital lobes, and insula along with the observation of grasping or
chewing behavior are associated with poor outcome {[}7{]}.

Despite the above findings, outside of multiple sclerosis, WMHs are not
routinely employed as a diagnostic measure in clinical practice. Their
presence within asymptomatic patients or in association with a variety
of conditions, such as stroke, dementia, neuroinflammatory conditions,
and TBI challenge their utility in narrowing a radiological differential
diagnosis. Further, performing a comprehensive manual counting of number
and distribution of lesions in the clinical setting is simply not
practical. Despite the limited inclusion of WMH observations in routine
radiological reports, two large meta-analyses demonstrated an
association between WMHs, cognitive function, increased risk of stroke,
dementia, and death {[}9, 10{]}. As such, the development of automated
methods for the rapid identification and quantification of WMHs within
individual patients may allow for identification of correlative patterns
between WMH number, volume, distribution, and disease state. Further,
the development of such lesion quantification approaches may allow for
the practical inclusion of this type of information within routine
radiological practice.

@@

Machine learning and pattern recognition techniques have seen increased
application for various medical image analysis workflows (see, for
example, the annual Workshop on Machine Learning in Medical Imaging held
in conjunction with the Medical Image Computing and Computer-Aided
Intervention (MICCAI) international meeting {[}11{]}). Popular
techniques such as support vector machines and neural networks have been
applied successfully to clinically relevant imaging tasks such as
supervised image segmentation (e.g., {[}12{]}) and diagnostic prediction
(e.g., {[}13, 14{]}). Facilitating the current employment of such
techniques are the number of available imaging data sets {[}15{]} and
the public availability of data science packages such as SciPy {[}16{]}
and the R project for statistical computing {[}17{]} and their
associated extensions.

The random forests framework {[}18{]} is a popular machine learning
technique that has demonstrated significant
\replaced[id=dt]{utility}{utitility} for supervised segmentation tasks
(e.g., normal human brain segmentation {[}19{]}) and other computer
vision applications (e.g., human gait detection {[}20{]}). In the
context of neuropathology, random forest-based paradigms have been
employed in the delineation of multiple sclerosis lesions {[}21{]},
stroke lesions {[}22{]}, and brain tumors {[}23--26{]}. Of note, these
latter random forest approaches for brain tumor segmentation have
performed well in recent international competitions
\replaced[id=dt]{established in response to the lack of objective comparisons between segmentation algorithms (i.e.,  the Multimodal Brain Tumor Segmentation (BRATS) challenge was initiated in 2012}{
In response to the lack of objective comparisons between segmentation algorithms, the Multimodal Brain Tumor Segmentation (BRATS) challenge was initiated in 2012 [@Menze:2015aa] and has continued every year since under the auspices of the MICCAI conference.}{[}27{]}.

Random forests are conceptually simple {[}18{]}. They consist of
ensembles of decision trees that are built from training data. Once
constructed, data to be classified is ``pushed'' through each decision
tree resulting in a single classification ``vote'' per tree. These votes
are then used for regression or classification of the data. Although
decision trees had been extensively studied \deleted[id=dt]{previously},
the success of employing collections of such weak learners for boosting
machine learning performance (e.g., AdaBoost {[}28, 29{]}) influenced
the similarly sytled conglomeration of decision trees into ``forests''
with randomized node optimization {[}30, 31{]}. Finally, Breiman
{[}18{]} improved accuracy by random sampling of \deleted[id=dt]{the}
training data (i.e., ``bagging'') resulting in the current random forest
technique \added[id=dt]{applied here}.

@@

Crucial to these supervised segmentation approaches are the creation and
selection of ``features'' as input in conjunction with
\replaced[id=lw,remark=NT: I replaced ground-truth since it is problematic anyway]
{expertly identified structures of interest}{the ground-truth} for model
construction. For the targeted application in this work (i.e., WMHs),
regression/classification are performed at the voxelwise level. In other
words, each voxel within the region of interest is sent through the
ensemble of decision trees and receives a set of classification votes
from each tree thus permitting a regression or classification solution.
Since this procedure is performed at the voxelwise level, intensity
information alone is insufficient for good segmentation performance due
to the lack of spatial context. For example, as pointed out in {[}32{]},
higher intensities can be found at the periventricular caps in normal
subjects which often confounds automated lesion detection algorithms.
Other potential confounds include MR signal inhomogeneity and noise.
Therefore, even though machine learning and pattern recognition
techniques are extremely powerful and have significant potential, just
as crucial to outcome is the creative construction and deployment of
salient feature images which we detail below.

@@

\begin{figure}[htbp]
\centering
\includegraphics{Figures/featureImages.png}
\caption{Representation of Stage 1 feature images for subject 01C1019.
The FLAIR, T1-, and T2-weighted images are
\added[id=am,remark=NT: no need for 12 dof for
intra-subject normalization]{rigidly} pre-aligned {[}33{]} to the space
of the T1 image. The three modality images are then preprocessed (N4
bias correction {[}34{]} and adaptive denoising {[}35{]}) followed by
application of standard ANTs brain extraction and \(n\)-tissue
segmentation protocols using the MMRR symmetric template and
corresponding priors {[}36{]} applied to the T1 image. The feature
images are then generated for voxelwise input to the RF model which
results in the voting maps illustrated on the right. This gives a
probabilistic classification of tissue type. Not shown are the
probability and voting images for the brain stem and cerebellum.}
\end{figure}

As mentioned previously, input for each subject comprises FLAIR, T1-,
and T2-weighted acquisitions. The FLAIR and T2 images are rigidly
registered to the T1 image using the open-source Advanced Normalization
Tools (ANTs) {[}33{]}. The aligned images are then preprocessed using
the denoising algorithm of {[}35{]} followed by N4 bias correction
{[}34{]} which are then normalized to the intensity range \([0,1]\).
Although we could have used an alternative intensity standardization
algorithm (e.g., {[}37{]}), we found that a simple linear rescaling
produced better results similar \added[id=dt]{to previous work}{[}26{]}.

\begin{figure}[htbp]
\centering
\includegraphics{Figures/montage.png}
\caption{Sample FLAIR acquisition image slices showing both manual (top)
and random forest (RDF) segmentations (bottom) obtained during the
leave-one-out evaluation. Manual segmentations were performed by one of
the authors and provided the ground truth WMH labels for training the
random forest models.}
\end{figure}

The T1 image is then processed via the ANTs brain extraction and normal
tissue segmentation (i.e., the pathology is discounted which is a fairly
safe assumption for the T1 image) protocols as part of the cortical
thickness pipeline described in {[}36{]}.
\added[id=hl]{This protocol involves preprocessing using N4 bias correction followed by a template-based strategy for brain extraction.  Once the
brain has been extracted, we apply a Bayesian-based segmentation algorithm
with a template-based prior probability strategy to segment the
parenchymal tissue types.  The result is} a mask delineating the brain
parenchyma and probabilistic estimates of the
\replaced[id=dt]{CSF}{cerebrospinal fluid (csf)}, gray matter, white
matter, deep gray matter, brain stem, and cerebellum. These provide the
expertly annotated labels for the first six tissue labels given above.
The \replaced[id=dt]{WMHs}{white matter hyperintensities} were manually
identified by one of the authors (J. R. S.) using the ITK-SNAP tool
{[}38{]}. Segmentation is performed using the ANTs Atropos tool {[}39{]}
and multi-model optimal symmetric shape/intensity templates {[}26{]}
created from the public MMRR data set {[}40{]} (cf Figure 3).

\begin{figure}[htbp]
\centering
\includegraphics{Figures/MMRR.png}
\caption{Canonical views of the mutlivariate, bilaterally symmetric
template constructed from the MMRR data set {[}40{]} (only shown are the
FLAIR, T1, and T2 modalities--- the components relevant for this work).
Template construction is detailed in {[}26{]}. These images are
important for specific intensity-based features.}
\end{figure}

To model the intensity information the first set of images simply
includes the preprocessed and normalized intensity FLAIR, T1, and T2
image voxel values. We also calculate a set of neighborhood statistics
(mean, standard deviation, and skewness) feature images using a
\added[id=am]{Manhattan} radius of one voxel
\added[id=nt]{given the typical size
of individual WMHs}. For each of the normalized images, we calculate the
difference in intensities with the corresponding warped template
component. Previous success in the international brain tumor
segmentation competition {[}27{]} was based on an important set of
intensity features that were created from multi-modal templates
mentioned previously {[}26{]} \added[id=nt]{and listed in Table 1}. We
employ the same strategy here. For example, the template difference
feature image for the FLAIR image, \(S_{FLAIR}\) is calculated as:

@@

The segmentation probability images described above are used as feature
images to provide a spatial context for the random forest model
prediction step. Additional spatial contextual feature images include
the distance maps {[}41{]} based on the csf, gray matter, and deep gray
matter images. These latter images are intended to help distinguish
white matter hyperintensities from false positives induced by the
partial voluming at the gray/white matter interface. A third set of
images are based on the voxel location within the space of the template.
The T1 image of the subject is registered to the T1 template component
using a B-spline variant {[}42{]} of the well-known ANTs Symmetric
Normalization (SyN) algorithm {[}43{]}. Since the inverse transform is
also derived as part of the registration process, we can warp the voxel
index locations back to the space of the individual subject
\replaced[id=nt]{which motivates similar work by others}{
Note that this is similar in motivation to the work
of} {[}44{]}. However, this previous work lacks the normalization to the
standard coordinate system provided by the template to dramatically
improve spatial specificity across all subjects.

@@

In previous brain tumor segmentation work {[}26{]}, it was demonstrated
that a concatenated supervised approach, whereby the prediction output
from the first random forest model serves as partial input for a second
random forest model, can significantly improve segmentation performance.
We do the same thing for the work described here
\added[id=lw]{where we employ two stacked random forests (or two "stages")}.
The Stage 1 feature images of the training data (as described
previously) are used to construct the Stage 1 model. The training data
Stage 1 features are then used to produce the voxelwise
\added[id=lw]{"voting maps" (i.e., the classification count of each decision tree for each tissue label)}
via the Stage 1 random forest model. All the Stage 1 features plus the
Stage 1 voting maps are used as input to the Stage 2 model. In addition,
we use the Stage 1 voting maps as tissue priors
\added[id=lw]{(i.e., probabilistic estimates of the tissue spatial locations)}
for a second application of the Atropos maximum aposteriori algorithm
with an additional Markov Random Field spatial prior (MAP-MRF) {[}39{]}.
However, for the second stage we use all three aligned preprocessed
images for a multivariate segmentation. The resulting seven posterior
probability images constitute a third additional feature image set for
Stage 2.

@@

As pointed out in a recent comprehensive multiple sclerosis lesion
segmentation review {[}45{]}, although the number of algorithms reported
in the literature is quite extensive, there were only four publicly
available segmentation algorithms at the time of writing
\added[id=am]{this article}.
\replaced[id=dt]{In contrast to the current work,}{of which} none are
based on supervised learning. As we did for our brain tumor segmentation
algorithm {[}26{]}, all of the code described in this work is publicly
available through the open-source ANTs/ANTsR toolkits. Through ANTsR (an
add-on toolkit which, in part, bridges ANTs and the R statistical
project) we use the \texttt{randomForest} package {[}46{]} using the
default settings with 2000 trees per model and 500 randomly selected
samples per label per image.
\added[id=bt]{Note that we saw little variation in performance when these parameters were changed which is consistent with our previous experience.}

@@

\begin{figure}[htbp]
\centering
\includegraphics{Figures/FLAIRcontralaleteralWithLesions.png}
\caption{\added[id=nt]{(a) FLAIR image slice illustrating WMHs which have been manually delineated
in (b).  (c) Contralateral FLAIR difference image.}}
\end{figure}

@@

In Figure 7 are the segmentation comparisons derived from manual
segmentations of the same data. Despite the large variability
characteristic with manual labelings in related fields {[}45, 47, 48{]},
such labelings are characteristic of current clinical practices and the
methodology proposed herein is readily adapted to refinements in
training data. On the left of Figure 7 are the improvement in Dice
values {[}49{]}, \added[id=nt]{i.e.,}

@@

@@

Two major meta-analyses of WMHs have been published covering the periods
of 1966-2009 and 2010-2015, respectively. The earlier meta-analysis
covered 53 longitudinal studies that included samples of high-risk
populations, i.e., patients selected for a specific disease or condition
such as hypertension, whereas other studies recruited samples of the
general population. Longitudinal studies of samples representative of
the general population are more relevant to the focus of the present
paper. Debette \& Markus {[}9{]} found that the presence of WMHs was
related to subsequent cognitive decline, a higher risk of developing
dementia, stroke, and of mortality. Lesion volume at baseline was also
predictive of cognitive decline. Limitations of this meta-analysis
include heterogeneity in the method of measuring WMHs; some studies used
automated volumetric measurement, whereas others used a visual rating
scale. The studies analyzed by Debette \& Markus were limited to the
occurrence of one of the aforementioned conditions which they analyzed
by hazard ratios.

>>>>

\usepackage{booktabs}
\usepackage[final]{changes}
\usepackage[font={small},labelfont=bf,labelsep=colon]{caption}
\linespread{1.5}
\usepackage[compact]{titlesec}
\usepackage{enumitem}
\usepackage{tikz}
\def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
\setlist{nolistsep}
\titlespacing{\section}{2pt}{*0}{*0}
\titlespacing{\subsection}{2pt}{*0}{*0}
\titlespacing{\subsubsection}{2pt}{*0}{*0}
\setlength{\parskip}{3pt}
\setremarkmarkup{(#2)}
\definechangesauthor[name={Nick Tustison}, color=red]{nt}
\definechangesauthor[name={James Stone}, color=magenta]{js}
\definechangesauthor[name={Lisa Wilde}, color=cyan]{lw}
\definechangesauthor[name={Andy Mayer}, color=green]{am}
\definechangesauthor[name={Harvey Levin}, color=blue]{hl}
\definechangesauthor[name={Brian Taylor}, color=purple]{bt}
\definechangesauthor[name={David Tate}, color=orange]{dt}

@@

\section{Abstract}\label{abstract}

White matter hyperintensities (WMHs) are foci of abnormal signal
intensity in white matter regions seen with magnetic resonance imaging
(MRI). WMHs are associated with normal aging and have shown prognostic
value in neurological conditions such as traumatic brain injury (TBI).
The impracticality of manually quantifying these lesions limits their
clinical utility and motivates the utilization of machine learning
techniques for automated segmentation workflows. Herein, we develop a
concatenated random forest framework with tailored features for
segmenting WMHs in a TBI cohort. The framework is provided publicly
through the Advanced Normalization Tools (ANTs) and ANTsR toolkits. MR
(3D FLAIR, T2-, and T1-weighted) images from 24 service members and
veterans scanned in the Chronic Effects of Neurotrauma Consortium's
(CENC) observational study were acquired. Manual annotations were
employed for both training and evaluation using a leave-one-out
strategy. Lesion load and overlap evaluative comparisons are
complimented by feature rankings which showcase the utility of the
concatenated approach. Our findings suggest supervised learning methods
may be applied to quantify WMHs on routine brain imaging. Paired with
correlative outcome data, supervised learning methods may allow for
identification of imaging features predictive of diagnosis and prognosis
in individual TBI patients.

@@

White matter hyperintensities (WMHs) are foci of abnormally increased
signal intensity seen within white matter regions within the cerebrum
and brainstem on fluid attenuation inversion recovery (FLAIR) magnetic
resonance imaging (MRI) sequences. These lesions are distinguished from
prominent perivascular spaces seen on T2-weighted imaging by the lack of
fluid suppression of signal on FLAIR sequences. WMHs in the
periventricular and deep brain regions are associated with normal aging
and neurological conditions including hypertension and stroke. WMHs are
also a frequent finding following traumatic brain injury (TBI) and have
been correlated with functional outcome and injury severity in both
pediatric\textsuperscript{1,2} and adult\textsuperscript{3--6}. Further,
the regional distribution and volume of WMHs have been shown to possess
prognostic value in the TBI patient\textsuperscript{2,6--8}.
Specifically, lesion volume in corpus callosum correlates with
functional scores in the acute phase following injury, while lesion
volume in frontal lobes correlates with scores at 1 year following
injury\textsuperscript{6}. Further, volume of FLAIR lesions within the
corpus callosum, brainstem, and thalamus in patients with severe TBI
correlates with Glasgow Outcome-Extended (GOS-E)
scores\textsuperscript{4}. Additionally, in patients who are comatose
following severe TBI the regional distribution of FLAIR lesions within
the pons, midbrain, hypothalamus, basal forebrain, parietal, temporal,
occipital lobes, and insula along with the observation of grasping or
chewing behavior are associated with poor outcome\textsuperscript{7}.

Despite the above findings, outside of multiple sclerosis, WMHs are not
routinely employed as a diagnostic measure in clinical practice. Their
presence within asymptomatic patients or in association with a variety
of conditions, such as stroke, dementia, neuroinflammatory conditions,
and TBI challenge their utility in narrowing a radiological differential
diagnosis. Further, performing a comprehensive manual counting of number
and distribution of lesions in the clinical setting is simply not
practical. Despite the limited inclusion of WMH observations in routine
radiological reports, two large meta-analyses demonstrated an
association between WMHs, cognitive function, increased risk of stroke,
dementia, and death\textsuperscript{9,10}. As such, the development of
automated methods for the rapid identification and quantification of
WMHs within individual patients may allow for identification of
correlative patterns between WMH number, volume, distribution, and
disease state. Further, the development of such lesion quantification
approaches may allow for the practical inclusion of this type of
information within routine radiological practice.

@@

Machine learning and pattern recognition techniques have seen increased
application for various medical image analysis workflows (see, for
example, the annual Workshop on Machine Learning in Medical Imaging held
in conjunction with the Medical Image Computing and Computer-Aided
Intervention (MICCAI) international meeting\textsuperscript{11}).
Popular techniques such as support vector machines and neural networks
have been applied successfully to clinically relevant imaging tasks such
as supervised image segmentation (e.g.,\textsuperscript{12}) and
diagnostic prediction (e.g.,\textsuperscript{13,14}). Facilitating the
current employment of such techniques are the number of available
imaging data sets\textsuperscript{15} and the public availability of
data science packages such as SciPy\textsuperscript{16} and the R
project for statistical computing\textsuperscript{17} and their
associated extensions.

The random forests framework\textsuperscript{18} is a popular machine
learning technique that has demonstrated significant
\replaced[id=dt]{utility}{utitility} for supervised segmentation tasks
(e.g., normal human brain segmentation\textsuperscript{19}) and other
computer vision applications (e.g., human gait
detection\textsuperscript{20}). In the context of neuropathology, random
forest-based paradigms have been employed in the delineation of multiple
sclerosis lesions\textsuperscript{21}, stroke
lesions\textsuperscript{22}, and brain tumors\textsuperscript{23--26}
for both single and multi-modal acquisition protocols. Of note, these
latter random forest approaches for brain tumor segmentation have
performed well in recent international competitions
\replaced[id=dt]{established in response to the lack of objective comparisons between segmentation algorithms (i.e.,  the Multimodal Brain Tumor Segmentation (BRATS) challenge was initiated in 2012}{
In response to the lack of objective comparisons between segmentation algorithms, the Multimodal Brain Tumor Segmentation (BRATS) challenge was initiated in 2012 [@Menze:2015aa] and has continued every year since under the auspices of the MICCAI conference.}\textsuperscript{27}.

Random forests are conceptually straightforward\textsuperscript{18}.
They consist of ensembles of decision trees that are built from training
data. Once constructed, data to be classified is ``pushed'' through each
decision tree resulting in a single classification ``vote'' per tree.
These votes are then used for regression or classification of the data.
Although decision trees had been extensively studied
\deleted[id=dt]{previously}, the success of employing collections of
such weak learners for boosting machine learning performance (e.g.,
AdaBoost\textsuperscript{28,29}) influenced the similarly sytled
conglomeration of decision trees into ``forests'' with randomized node
optimization\textsuperscript{30,31}. Finally,
Breiman\textsuperscript{18} improved accuracy by random sampling of
\deleted[id=dt]{the} training data (i.e., ``bagging'') resulting in the
current random forest technique \added[id=dt]{applied here}.

@@

Crucial to these supervised segmentation approaches are the creation and
selection of ``features'' as input in conjunction with
\replaced[id=lw,remark=NT: I replaced ground-truth since it is problematic anyway]
{expertly identified structures of interest}{the ground-truth} for model
construction. For the targeted application in this work (i.e., WMHs),
regression/classification are performed at the voxelwise level. In other
words, each voxel within the region of interest is sent through the
ensemble of decision trees and receives a set of classification votes
from each tree thus permitting a regression or classification solution.
Since this procedure is performed at the voxelwise level, intensity
information alone is insufficient for good segmentation performance due
to the lack of spatial context. For example, as pointed out
in\textsuperscript{32}, higher intensities can be found at the
periventricular caps in normal subjects which often confounds automated
lesion detection algorithms. Other potential confounds include MR signal
inhomogeneity and noise. Therefore, even though machine learning and
pattern recognition techniques are extremely powerful and have
significant potential, just as crucial to outcome is the creative
construction and deployment of salient feature images which we detail
below.

@@

\begin{figure}[htbp]
\centering
\includegraphics{Figures/featureImages.png}
\caption{Representation of Stage 1 feature images for subject 01C1019.
The FLAIR, T1-, and T2-weighted images are
\added[id=am,remark=NT: no need for 12 dof for
intra-subject normalization]{rigidly} pre-aligned\textsuperscript{33} to
the space of the T1 image. The three modality images are then
preprocessed (N4 bias correction\textsuperscript{34} and adaptive
denoising\textsuperscript{35}) followed by application of standard ANTs
brain extraction and \(n\)-tissue segmentation protocols using the MMRR
symmetric template and corresponding priors\textsuperscript{36} applied
to the T1 image. The feature images are then generated for voxelwise
input to the RF model which results in the voting maps illustrated on
the right. This gives a probabilistic classification of tissue type. Not
shown are the probability and voting images for the brain stem and
cerebellum.}
\end{figure}

As mentioned previously, input for each subject comprises FLAIR, T1-,
and T2-weighted acquisitions. The FLAIR and T2 images are rigidly
registered to the T1 image using the open-source Advanced Normalization
Tools (ANTs)\textsuperscript{33}. The aligned images are then
preprocessed using the denoising algorithm of\textsuperscript{35}
followed by N4 bias correction\textsuperscript{34} which are then
normalized to the intensity range \([0,1]\). Although we could have used
an alternative intensity standardization algorithm
(e.g.,\textsuperscript{37}), we found that a simple linear rescaling
produced better results similar
\added[id=dt]{to previous work}\textsuperscript{26}.

\begin{figure}[htbp]
\centering
\includegraphics{Figures/sampleResults.png}
\caption{Sample FLAIR acquisition image slices showing both manual and
random forest segmentations for both stages obtained during the
leave-one-out evaluation. Manual segmentations were performed by one of
the authors and provided the ground truth WMH labels for training the
random forest models.}
\end{figure}

The T1 image is then processed via the ANTs brain extraction and normal
tissue segmentation (i.e., the pathology is discounted which is a fairly
safe assumption for the T1 image) protocols as part of the cortical
thickness pipeline described in\textsuperscript{36}.
\added[id=hl]{This protocol involves preprocessing using N4 bias correction followed by a template-based strategy for brain extraction.  Once the
brain has been extracted, we apply a Bayesian-based segmentation algorithm
with a template-based prior probability strategy to segment the
parenchymal tissue types.  The result is} a mask delineating the brain
parenchyma and probabilistic estimates of the
\replaced[id=dt]{CSF}{cerebrospinal fluid (csf)}, gray matter, white
matter, deep gray matter, brain stem, and cerebellum. These provide the
expertly annotated labels for the first six tissue labels given above.
The \replaced[id=dt]{WMHs}{white matter hyperintensities} were manually
identified by one of the authors (J. R. S.) using the ITK-SNAP
tool\textsuperscript{38}. Segmentation is performed using the ANTs
Atropos tool\textsuperscript{39} and multi-model optimal symmetric
shape/intensity templates\textsuperscript{26} created from the public
MMRR data set\textsuperscript{40} (cf Figure 3).

\begin{figure}[htbp]
\centering
\includegraphics{Figures/MMRR.png}
\caption{Canonical views of the mutlivariate, bilaterally symmetric
template constructed from the MMRR data set\textsuperscript{40} (only
shown are the FLAIR, T1, and T2 modalities--- the components relevant
for this work). Template construction is detailed
in\textsuperscript{26}. These images are important for specific
intensity-based features.}
\end{figure}

To model the intensity information the first set of images simply
includes the preprocessed and normalized intensity FLAIR, T1, and T2
image voxel values. We also calculate a set of neighborhood statistics
(mean, standard deviation, and skewness) feature images using a
\added[id=am]{Manhattan} radius of one voxel
\added[id=nt]{given the typical size
of individual WMHs}. For each of the normalized images, we calculate the
difference in intensities with the corresponding warped template
component. Previous success in the international brain tumor
segmentation competition\textsuperscript{27} was based on an important
set of intensity features that were created from multi-modal templates
mentioned previously\textsuperscript{26}
\added[id=nt]{and listed in Table 1}. We employ the same strategy here.
For example, the template difference feature image for the FLAIR image,
\(S_{FLAIR}\) is calculated as:

@@

The segmentation probability images described above are used as feature
images to provide a spatial context for the random forest model
prediction step. Additional spatial contextual feature images include
the distance maps\textsuperscript{41} based on the csf, gray matter, and
deep gray matter images. These latter images are intended to help
distinguish white matter hyperintensities from false positives induced
by the partial voluming at the gray/white matter interface. A third set
of images are based on the voxel location within the space of the
template. The T1 image of the subject is registered to the T1 template
component using a B-spline variant\textsuperscript{42} of the well-known
ANTs Symmetric Normalization (SyN) algorithm\textsuperscript{43}. Since
the inverse transform is also derived as part of the registration
process, we can warp the voxel index locations back to the space of the
individual subject
\replaced[id=nt]{which motivates similar work by others}{
Note that this is similar in motivation to the work
of}\textsuperscript{44}. However, this previous work lacks the
normalization to the standard coordinate system provided by the template
to dramatically improve spatial specificity across all subjects.

@@

In previous brain tumor segmentation work\textsuperscript{26}, it was
demonstrated that a concatenated supervised approach, whereby the
prediction output from the first random forest model serves as partial
input for a second random forest model, can significantly improve
segmentation performance. We do the same thing for the work described
here
\added[id=lw]{where we employ two stacked random forests (or two "stages")}.
The Stage 1 feature images of the training data (as described
previously) are used to construct the Stage 1 model. The training data
Stage 1 features are then used to produce the voxelwise
\added[id=lw]{"voting maps" (i.e., the classification count of each decision tree for each tissue label)}
via the Stage 1 random forest model. All the Stage 1 features plus the
Stage 1 voting maps are used as input to the Stage 2 model. In addition,
we use the Stage 1 voting maps as tissue priors
\added[id=lw]{(i.e., probabilistic estimates of the tissue spatial locations)}
for a second application of the Atropos maximum aposteriori algorithm
with an additional Markov Random Field spatial prior
(MAP-MRF)\textsuperscript{39}. However, for the second stage we use all
three aligned preprocessed images for a multivariate segmentation. The
resulting seven posterior probability images constitute a third
additional feature image set for Stage 2.

@@

As pointed out in a recent comprehensive multiple sclerosis lesion
segmentation review\textsuperscript{45}, although the number of
algorithms reported in the literature is quite extensive, there were
only four publicly available segmentation algorithms at the time of
writing \added[id=am]{this article}.
\replaced[id=dt]{In contrast to the current work,}{of which} none are
based on supervised learning. As we did for our brain tumor segmentation
algorithm\textsuperscript{26}, all of the code described in this work is
publicly available through the open-source ANTs/ANTsR toolkits. Through
ANTsR (an add-on toolkit which, in part, bridges ANTs and the R
statistical project) we use the \texttt{randomForest}
package\textsuperscript{46} using the default settings with 2000 trees
per model and 500 randomly selected samples per label per image.
\added[id=bt]{Note that we saw little variation in performance when these parameters were changed (i.e. up to 1000 random samples and as little as 1000 trees) which is consistent with our previous experience.}

@@

\begin{figure}[htbp]
\centering
\includegraphics{Figures/FLAIRcontralaleteralWithLesionsAlternate.png}
\caption{\added[id=nt]{(a) FLAIR image slice illustrating WMHs which have been manually delineated.
The region around the WMHs is enlarged (b) in the original FLAIR and the
(c) contralateral FLAIR difference image.}}
\end{figure}

@@

In Figure 7 are the segmentation comparisons derived from manual
segmentations of the same data. Despite the large variability
characteristic with manual labelings in related
fields\textsuperscript{45,47,48}, such labelings are characteristic of
current clinical practices and the methodology proposed herein is
readily adapted to refinements in training data. On the left of Figure 7
are the improvement in Dice values\textsuperscript{49},
\added[id=nt]{i.e.,}

@@

\section{}\label{section}

@@

Two major meta-analyses of WMHs have been published covering the periods
prior to 2010\textsuperscript{9} and after\textsuperscript{10}. The
earlier meta-analysis covered 53 longitudinal studies that included
samples of high-risk populations, i.e., patients selected for a specific
disease or condition such as hypertension, whereas other studies
recruited samples of the general population. Longitudinal studies of
samples representative of the general population are more relevant to
the focus of the present paper. Debette \& Markus\textsuperscript{9}
found that the presence of WMHs was related to subsequent cognitive
decline, a higher risk of developing dementia, stroke, and of mortality.
Lesion volume at baseline was also predictive of cognitive decline.
Limitations of this meta-analysis include heterogeneity in the method of
measuring WMHs; some studies used automated volumetric measurement,
whereas others used a visual rating scale. The studies analyzed by
Debette \& Markus were limited to the occurrence of one of the
aforementioned conditions which they analyzed by hazard ratios.

\replaced[id=dt]{The more}{A} recent meta-analysis by Kloppenborg et
al.\textsuperscript{10} of 23 cross-sectional studies reporting MRI and
concurrent neuropsychological results in patients with heterogeneous
diagnoses but without previously diagnosed cognitive impairment, found
that WMHs were associated with cognitive deficit (effect size of -0.10,
95\% CI: -0.13 to -0.08) after controlling for age. These studies also
differed in the metric used to measure the WMHs, including volume, \% of
total intracranial volume, and a visual rating score. The effect size
for the association with cognitive deficit in these cross-sectional
studies did not differ significantly across various cognitive domains or
the method of measuring lesion volume. Among eight longitudinal studies
analyzed by Kloppenborg et al that included a follow-up MRI and also
controlled for age, the effect size for the association of progression
in WMHs and cognitive impairment was -0.16 (95\% CI:-0.27 to -0.09).
This association was stronger for attention and executive function than
for memory and processing speed. Although baseline WMHs were predictive
of cognitive deficit at follow-up in the seven studies which did not
repeat MRI, the effect size was smaller {[}-0.10 (95\% CI: 0.13 to
-0.05) than in the longitudinal studies that calculated progression in
WMHs. In summary, progression of WMHs seen on repeat MRI has a stronger
relation to cognitive deficit than concurrent imaging findings. These
meta-analyses support the rationale for repeating an MRI in patients
younger than 50 years whose initial scan shows WMHs.

Despite the above-described associations between WMHs, cognitive
decline, increased risk of developing dementia, and mortality, these
lesions receive little attention in current clinical workflows. When
reported in a standard neuroradiologist interpretation, they are
typically handled as incidental findings and are assigned little
clinical significance. This likely reflects the impracticality of
performing a detailed assessment of number, volume, and distribution
within a qualitative neuroradiologist interpretation as well as the lack
of correlative information on how the presence and distribution of these
lesions may inform a diagnosis and prognosis in the appropriate clinical
setting. To date, automated or semi-automated tools for the detection of
WMHs have lacked the specificity and efficiency for the mining of
large-scale datasets to generate highly granular data on whether these
lesions possess any true diagnostic or prognostic value in the setting
of a specific disease process. The present communication describes a
supervised statistical learning tool that is appropriate for the
application to such large-scale datasets.
